Shapes analysis for time series.

### 2024-05-07 NEURIPS24/neurips_2024_to_submit.tex
update

\begin{abstract}
    %TO CHANGE]
    %Analyzing the inter-individual variability within a cluster of time series is particularly appealing in medicine and biology.
    %Individuals have specificities depending on their genes or disease
     %while sharing, on average, a similar pattern in the features of their corresponding time series. For instance, an electrocardiogram presents a typical shape of the heartbeat, which varies depending on the individual.
    %In this paper, we propose an unsupervised representation method (TS-LDDMM) to analyze inter-individual variability of time series dataset.
    %Inspired by the shape analysis literature, we extend Large Deformation Diffeomorphic Metric Mapping (LDDMM) to time series, considering the temporal evolution as their distinctive feature.  
    % Given a dataset of $N$ individual's time series $(s^j)_{j\in[N]}$,
    %  we learn a time series of reference $\mathbf{s}_0$ encoding the common part of the individuals as well as diffeomorphisms $(\phi^j)_{j\in[N]}$ encoding the specificities of each individual:
    %  $s^j$ is seen as the transformation of $\mathbf{s}_0$ by $\phi_j$.
    %  Then, the parameters $(\alpha_j)_{j\in[N]}$ encoding the diffeomorphisms $(\phi^j)_{j\in[N]}$ are used as the features representation related to each time series $s^j$.
    In this paper, we propose an unsupervised representation learning method (URL) for time series %\cite{trirat2024universal,meng2023unsupervised}
     by giving a special attention to their \textit{shapes} for studying inter-variability in biomedical applications.
    In particular, the method can handle \textit{irregularly sampled} time series having \textit{different sizes}.
    This work belongs to the shape analysis literrature: %which focus on the statistical analysis of various mathematical objects invariant under rotations, dilations, or time parameterization.
     we extend Large Deformation Diffeomorphic Metric Mapping (LDDMM) %\cite{beg2005computing,vaillant2004statistics}
      to the case of time series data by applying deformations to their graph.
    However, the method is not a simple application of LDDMM since time series graph have more structure than 1D curves.
     Indeed, a graph time series transformed by a general diffeomorphism is not always a graph time series.
      We solve this issue by establishing a representation theorem on time series graph and derive its consequences on LDDMM.
    This work is at the crossroad of URL for time series and shape analysis; we hope it will benefit to both communities.
    % These features can be post-processed with any statistical or machine learning tools. 
     We demonstrate the advantages of our representation compared to existing methods using synthetic data and real-world examples, motivated by applications in medicine.
\end{abstract}

>>>>

\begin{abstract}
    %TO CHANGE]
    %Analyzing the inter-individual variability within a cluster of time series is particularly appealing in medicine and biology.
    %Individuals have specificities depending on their genes or disease
     %while sharing, on average, a similar pattern in the features of their corresponding time series. For instance, an electrocardiogram presents a typical shape of the heartbeat, which varies depending on the individual.
    %In this paper, we propose an unsupervised representation method (TS-LDDMM) to analyze inter-individual variability of time series dataset.
    %Inspired by the shape analysis literature, we extend Large Deformation Diffeomorphic Metric Mapping (LDDMM) to time series, considering the temporal evolution as their distinctive feature.  
    % Given a dataset of $N$ individual's time series $(s^j)_{j\in[N]}$,
    %  we learn a time series of reference $\mathbf{s}_0$ encoding the common part of the individuals as well as diffeomorphisms $(\phi^j)_{j\in[N]}$ encoding the specificities of each individual:
    %  $s^j$ is seen as the transformation of $\mathbf{s}_0$ by $\phi_j$.
    %  Then, the parameters $(\alpha_j)_{j\in[N]}$ encoding the diffeomorphisms $(\phi^j)_{j\in[N]}$ are used as the features representation related to each time series $s^j$.
    In this paper, we propose an unsupervised representation learning method (URL) for time series %\cite{trirat2024universal,meng2023unsupervised}
     by giving a special attention to their \textit{shapes} for studying inter-variability in biomedical applications.
    In particular, the method can handle \textit{irregularly sampled} time series with \textit{variable sizes}.
    This work belongs to the shape analysis literrature: %which focus on the statistical analysis of various mathematical objects invariant under rotations, dilations, or time parameterization.
     we extend Large Deformation Diffeomorphic Metric Mapping (LDDMM) %\cite{beg2005computing,vaillant2004statistics}
      to the case of time series data by applying deformations to their graph.
    However, the method is not a simple application of LDDMM since time series graph have more structure than curves.
     Indeed, a graph time series transformed by a general diffeomorphism is not always a graph time series.
      We solve this issue by establishing a representation theorem on time series graph and derive its consequences on LDDMM.
    This work is at the crossroad of URL for time series and shape analysis; we hope it will benefit to both communities.
    % These features can be post-processed with any statistical or machine learning tools. 
     We demonstrate the advantages of our representation compared to existing methods using synthetic data and real-world examples, motivated by applications in medicine.
\end{abstract}

### 2024-05-07 NEURIPS24/paper.tex
update

   However, a graph time series transformation by a general diffeomorphism is not always a graph time series, see e.g. \Cref{fig:diffeo}, thus a graph time series is more than a simple 1D curve \cite{glaunes2008large}.
   Our contributions arise from this observation.
   %In this regard, this work is not an application of LDDMM to 1D curves \cite{glaunes2008large}.
    We solve this issue by specifying the class of diffeomorphisms to consider and showing how to learn them.
     This change is fruitful in representing time transformation as illustrated in \Cref{fig:transport}.
     In particular, the method can handle \textit{irregularly sampled} time series having \textit{different sizes}.

     Our contributions can be summarized as follows:
     \begin{itemize}
       \item We propose an unsupervised method (TS-LDDMM) to analyze inter-individual variability of shapes in a time series dataset.
       We especially motivate our extension of LDDMM to time series by introducing a theoretical framework with a representation theorem for time series graph (\Cref{theorem:representation}) and kernels related to their structure (\Cref{lemma:choice_of_kernel_V}).
       \item We demonstrate the identifiability of the model by estimating the true generating parameter of synthetic data, and we highlight the sensitivity of our method with respect to its hyperparameters, also providing guidelines for tuning.
       We highlight the \textit{interpretability} of TS-LDDMM for studying the inter-individual variability in a clinical dataset.
        We illustrate the quantitative interest of the representation on classification tasks on real shape-based datasets.

>>>>

   However, a graph time series transformation by a general diffeomorphism is not always a graph time series, see e.g. \Cref{fig:diffeo}, thus a graph time series is more than a simple curve \cite{glaunes2008large}.
   Our contributions arise from this observation.
   %In this regard, this work is not an application of LDDMM to 1D curves \cite{glaunes2008large}.
    We solve this issue by specifying the class of diffeomorphisms to consider and showing how to learn them.
     This change is fruitful in representing time transformation as illustrated in \Cref{fig:transport}.
     %In particular, the method can handle \textit{irregularly sampled} time series with \textit{variable sizes}.

     Our contributions can be summarized as follows:
     \begin{itemize}
       \item We propose an unsupervised method (TS-LDDMM) to analyze inter-individual variability of shapes in a time series dataset. In particular, the method can handle \textit{irregularly sampled} time series with \textit{variable sizes}.
       \item We motivate our extension of LDDMM to time series by introducing a theoretical framework with a representation theorem for time series graph (\Cref{theorem:representation}) and kernels related to their structure (\Cref{lemma:choice_of_kernel_V}).
       \item We demonstrate the identifiability of the model by estimating the true generating parameter of synthetic data, and we highlight the sensitivity of our method with respect to its hyperparameters, also providing guidelines for tuning.
       We highlight the \textit{interpretability} of TS-LDDMM for studying the inter-individual variability in a clinical dataset.
        We illustrate the quantitative interest of the representation on classification tasks on real shape-based datasets.
